### 2016-02-09 mcmc.tex
problems, and conversions of HOGG to DFM

The autocorrelation function is defined by
\begin{eqnarray}
C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
\label{eq:autocorrelationfn}\\
\mu &=& E[\pars]
\quad ;
\end{eqnarray}
that is, it is the covariance of points separated by $\Delta$, taken
relative to the variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau = \sum_{\Delta = 0}^{\infty} C(\Delta)
\label{eq:autocorrelationtime}\quad .
\end{eqnarray}
%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING? OR A SUM FROM -infty to +infty?
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

@@

A sampler with a smaller autocorrelation time is just better; you have
to do fewer $f(\pars)$ calls per independent sample, and you have to
run less time to get accurate sampling-based integral estimates.
A sampler that takes an independent sample every time would have an
autocorrelation time of unity, which is the best possible value; this
optimal sampling is only possible for problems where the sampling is
analytic (for example if the posterior is perfectly Gaussian with
known mean and variance).
In general, the best sampler will be different for different problems,
and we will (below; \sectionname~\ref{sec:methods}) tune the samplers we have to do the best on the
problems we have; this tuning will also be problem-specific.
There are many heuristic bases on which different samplers might be
compared or tuned, but fundamentally it is lower autocorrelation time
that separates good samplers from bad ones and is the ultimate basis
on which we compare performance.\note{%
  We sometimes see \MCMC\ methods compared according to burn-in times,
  which, for one, depends strongly on initialization, and, for two,
  depends strongly on tuning and dynamics.  Similarly we often see
  comparisons in terms of acceptance ratio.  In principle the only
  question is how precise are one's inferences given a certain amount
  of computation.  This is set by the autocorrelation time and (pretty
  much) nothing else, for reasonably converged chains.  In the real
  world, of course, one must add to the CPU time the investigator time
  spent tuning the method (and thinking about initialization) but
  there is never (to our knowledge) a condition in which burn-in time
  is the dominant consideration in choosing an MCMC method.}

@@

\begin{problem}
Write a piece of code that computes the empirical autocorrelation
function and the empirical autocorrelation time from a chain.  Run
this on the chain you obtained from \problemname~\ref{prob:MH}.
Plot the autocorrelation function you get and over-plot the
autocorrelation time as an indicator on the plot.  What do you
conclude?
\end{problem}

@@

\begin{problem}
Look at the source code of HOGG: REFERENCE EMCEE, and see if you can
re-code or extract the autocorrelation-time estimator called HOGG.
This code should take a \MCMC\ chain of parameter vectors as input and
return a vector of autocorrelation times as output.
\end{problem}

@@

\begin{problem}
Re-do \problemname~\ref{prob:tuning} but instead of trying to reach a
certain acceptance fraction, try to minimize the autocorrelation time.
(This, by the way, is the Right Thing To Do, but often expensive.)
What do you get as the best value of $Q$ in this case?
Again, just evaluate on a coarse logarithmic grid.
\end{problem}

@@

Just as most (though not all) \MCMC\ methods require a choice of
proposal distribution, most (though not all) require a choice about
initialization:
The walker needs (or walkers need) to be started somewhere in the
parameter space.
In many use cases for \MCMC, the investigator wants to use it to obtain
uncertainty information about or propagate uncertainty into parameter
estimates.
In these cases, it makes sense to initialize the walker or walkers at
sensible parameter estimates, found by optimizing a likelihood or a
posterior PDF in advance of sampling.
In extremely high dimensions (large numbers of parameters), this can
be a bad idea, since the optimal parameters are not necessarily near
typical posterior samples\note{When the dimensionality of the
  parameter space gets very large, almost all samples will be
  \emph{very far} from the optimum of the posterior pdf.  One way to
  think about this is to think of a random Gaussian draw in $D$
  dimensions: A typical draw will be $\sqrt{D}$ standard deviations
  Euclidean distance from the maximum of the Gaussian.  That's a long
  way when $D=100$ or $1000$!  Another way to think about it is that
  there are a lot of ways to move \emph{away} from the point you care
  about, but very few ways to move close to it.  The upshot is that
  even with an \emph{enormous} sampling, if you are in large
  dimensions, not a single sample will be very near the optimum.}, but
in low dimensions (few to tens) this is often sensible.

@@

As we mentioned above, scientists often like to report the ``best sample'';
  that is, the sample with the highest posterior pdf value.
This is not usually a good idea.
For one, \MCMC\ is a sampling algorithm, not an optimization algorithm;
  there are no guarantees that it will find the optimum of the posterior pdf in reasonable time.%
  \note{The sampler samples the pdf fairly.
    The tallest peak in the posterior is not guaranteed to be---%
    and in general is not---%
    also the peak with the greatest posterior ``mass''.
    That is, the samples in a fair sampling will not necessarily come
    predominantly from the \emph{tallest} peak in the posterior pdf.
    They will come from the peak with the greatest total integral
    of the likelihood over the prior.
    These issues might sound ``academic'',
    but when the number of parameters gets large (greater than, say, 10),
    many normal intuitions one might have (if any) about what is reasonably likely in a pdf
    are regularly violated.}
For two, the closeness of the best sample to the posterior mode is a strong function of sample size,
  and with very bad scaling.%
  \note{If you have a $K$-point sampling in a $D$-dimensional parameter space,
    there is some best sample $\pars_k^{\best}$.  Now imagine that you want to make
    that best sample 10 times better---that is, you want to find a point $\pars^{\better}$
    that is ten times closer to the true optimum.  How many more samples do you need to take?
    In the limit that $K$ is large, on average, you have to take a factor of
    $10^D$ more samples!
    Any reasonable optimizer is far, far better than this (in terms of compute time).}
For three, if you just want an optimum, use an optimizer!
That is, we \emph{strongly advise against} reporting,
  as ``the'' measurement or ``the result'',
  the parameter value or values for the best sample.
Giving some posterior samples is a good idea (we return to this below).
If you feel drawn to give the best sample,
  instead give the parameter values found by optimizing the posterior PDF,
  starting at the best sample as an initialization
  (and sacrifice your probabilism\note{something}).
This optimal-posterior parameter value is called the
  ``maximum \foreign{a posteriori}'' or ``MAP'' value for the parameters;
  it is like a maximum-likelihood value but regularized by the prior.
It is an estimator with some good (and some bad) properties, but it is not the point of \MCMC,
  and therefore outside the scope of this \documentname.

@@

There is one oddity to note here, as it catches many investigators by
surprise:
Even if the model is good and the posterior pdf is unimodal and well-behaved,
the median (or even mean) of the posterior pdf for each of parameters,
when taken together as a ``best-fit'' parameter vector $\pars^\ast$,
will not itself necessarily be a good fit to the data!
If there is substantial ``curvature'' to the pdf in the parameter
space, the mean or median of sampling does not necessarily lie at a
high-probability location of parameter space.
This may seem counterintuitive, but it is easy to see in the case of a
``banana-shaped'' posterior pdf.
This all relates to the fact that \MCMC\ is not an optimizer, it is a
sampler.
It is important, when reporting the output of an \MCMC\ run by giving
means or medians of the posterior PDF to remind the reader or user of
that output that it does not necessarily represent (collectively) a
good fit to the data; these outputs only give information that is
useful one parameter at a time.

@@

\paragraph{initialization-dependence:}
If you initialize your \MCMC\ runs in different places in parameter space,
  do you get the same (or very similar) final parameter samplings,
  in mean and variance?
If you don't,
  then your posterior PDF is badly multi-modal;
  trivial \MCMC\ methods probably won't fix your problem.
Not to put too fine a point on it:
If your results are initialization-dependent,
  then your \MCMC\ runs \emph{are not converged}.

You either have to go to a method that tries to explore the posterior
  more liberally, such as nested sampling or simulated tempering,
  or else split your ``model'' up into several sub-models,
  each of which contains different posterior modes.
In general, this problem is hard to fix;
  as we mentioned above, it is provably impossible
  to explore all of parameter space if the posterior PDF is
  complex in morphology.
We don't have very useful advice here, except to
  spend time learning about the multiple modes in the posterior PDF
  and what each ``means'' or corresponds to,
  and to do your best to express in your results the multiple optima.
One of the miracles of \MCMC\ is that if you \emph{can} fairly sample
  as complex posterior PDF,
  the fraction of samples in each mode
  tends to the relative total integrated posterior probability inside each mode,
  as the sampling converges.

@@

\paragraph{parameterization problems:}
It helps to think carefully about how to parameterize.
For example, if there is an angle $\phi$ in your problem,
  and your prior requires it to be between $0$ and $2\pi$,
  and the posterior mode is very near $0$,
  a little bit of probability leaking below $0$ plus the prior cutoffs plus mod-$2\pi$ symmetry
  makes an intrinsically unimodal problem multimodal!
In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
  to two vector components $a\equiv A\,\cos\phi$ and $b\equiv A\,\sin\phi$.
Of course this re-parameterization changes significantly the \emph{form} of the prior PDF
  (the transformation brings in a Jacobian);
  it has to be adjusted with care.

>>>>

The autocorrelation function is defined by
\begin{eqnarray}
C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
\label{eq:autocorrelationfn}\\
\mu &=& E[\pars]
\quad ;
\end{eqnarray}
that is, it is the covariance of points separated by $\Delta$, taken
relative to the variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau = 2\,\sum_{\Delta = 0}^{\infty} C(\Delta)
\label{eq:autocorrelationtime}\quad .
\end{eqnarray}
%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING? OR A SUM FROM -infty to +infty?
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

DFM: Here or in a note, can you give an empirical estimator of the
autocorrelation time?  The naive estimator is good to give, and we
could also give a good estimator too.

@@

A sampler with a smaller autocorrelation time is just better; you have
to do fewer $f(\pars)$ calls per independent sample, and you have to
run less time to get accurate sampling-based integral estimates.
A sampler that takes an independent sample every time would have an
autocorrelation time of unity, which is the best possible value; this
optimal sampling is only possible for problems where the sampling is
analytic (for example if the posterior is perfectly Gaussian with
known mean and variance).
In general, the best sampler will be different for different problems,
and we will (below; \sectionname~\ref{sec:methods}) tune the samplers we have to do the best on the
problems we have; this tuning will also be problem-specific.
There are many heuristic bases on which different samplers might be
compared or tuned, but fundamentally it is lower autocorrelation time
that separates good samplers from bad ones and is the ultimate basis
on which we compare performance.\note{%
  We sometimes see \MCMC\ methods compared according to burn-in times,
  which, for one, depends strongly on initialization, and, for two,
  depends strongly on tuning and dynamics.  Similarly we often see
  comparisons in terms of acceptance ratio.  In principle the only
  question is how precise are one's inferences given a certain amount
  of computation.  This is set by the autocorrelation time and (pretty
  much) nothing else, for reasonably converged chains.  In the real
  world, of course, one must add to the CPU time the investigator time
  spent tuning the method (and thinking about initialization) but
  there is never (to our knowledge) a condition in which burn-in time
  is the dominant consideration in choosing an \MCMC\ method.}

@@

\begin{problem}\label{prob:estimatetau}
Write a piece of code that computes the empirical autocorrelation
function and the empirical autocorrelation time from a chain using the na\"ive method DFM:WHERE?  Run
this on the chain you obtained from \problemname~\ref{prob:MH}.
Plot the autocorrelation function you get and over-plot the
autocorrelation time as an indicator on the plot.  What do you
conclude?
\end{problem}

\begin{problem}
Look at the source code of \project{emcee} \citep{emcee}, and see if you can
re-code or extract the autocorrelation-time estimator called DFM:WHAT?
This code should take a \MCMC\ chain of parameter vectors as input and
return a vector of autocorrelation times as output.
\end{problem}

@@

@@

\begin{problem}
Re-do \problemname~\ref{prob:tuning} but instead of trying to reach a
certain acceptance fraction, try to minimize the autocorrelation time.
You will need one of the autocorrelation-time estimators you might have
built in a previous \problemname.
(This, by the way, is the Right Thing To Do, but often expensive.)
What do you get as the best value of $Q$ in this case?
Again, just evaluate on a coarse logarithmic grid.
\end{problem}

@@

Just as most (though not all) \MCMC\ methods require a choice of
proposal distribution, most (though not all) require a choice about
initialization:
The walker needs (or walkers need) to be started somewhere in the
parameter space.
In many use cases for \MCMC, the investigator wants to use it to obtain
uncertainty information about or propagate uncertainty into parameter
estimates.
In these cases, it makes sense to initialize the walker or walkers at
sensible parameter estimates, found by optimizing a likelihood or a
posterior pdf in advance of sampling.
In extremely high dimensions (large numbers of parameters), this can
be a bad idea, since the optimal parameters are not necessarily near
typical posterior samples\note{When the dimensionality of the
  parameter space gets very large, almost all samples will be
  \emph{very far} from the optimum of the posterior pdf.  One way to
  think about this is to think of a random Gaussian draw in $D$
  dimensions: A typical draw will be $\sqrt{D}$ standard deviations
  Euclidean distance from the maximum of the Gaussian.  That's a long
  way when $D=100$ or $1000$!  Another way to think about it is that
  there are a lot of ways to move \emph{away} from the point you care
  about, but very few ways to move close to it.  The upshot is that
  even with an \emph{enormous} sampling, if you are in large
  dimensions, not a single sample will be very near the optimum.}, but
in low dimensions (few to tens) this is often sensible.

@@

As we mentioned above, scientists often like to report the ``best sample'';
  that is, the sample with the highest posterior pdf value.
This is not usually a good idea.
For one, \MCMC\ is a sampling algorithm, not an optimization algorithm;
  there are no guarantees that it will find the optimum of the posterior pdf in reasonable time.%
  \note{The sampler samples the pdf fairly.
    The tallest peak in the posterior is not guaranteed to be---%
    and in general is not---%
    also the peak with the greatest posterior ``mass''.
    That is, the samples in a fair sampling will not necessarily come
    predominantly from the \emph{tallest} peak in the posterior pdf.
    They will come from the peak with the greatest total integral
    of the likelihood over the prior.
    These issues might sound ``academic'',
    but when the number of parameters gets large (greater than, say, 10),
    many normal intuitions one might have (if any) about what is reasonably likely in a pdf
    are regularly violated.}
For two, the closeness of the best sample to the posterior mode is a strong function of sample size,
  and with very bad scaling.%
  \note{If you have a $K$-point sampling in a $D$-dimensional parameter space,
    there is some best sample $\pars_k^{\best}$.  Now imagine that you want to make
    that best sample 10 times better---that is, you want to find a point $\pars^{\better}$
    that is ten times closer to the true optimum.  How many more samples do you need to take?
    In the limit that $K$ is large, on average, you have to take a factor of
    $10^D$ more samples!
    Any reasonable optimizer is far, far better than this (in terms of compute time).}
For three, if you just want an optimum, use an optimizer!
That is, we \emph{strongly advise against} reporting,
  as ``the'' measurement or ``the result'',
  the parameter value or values for the best sample.
Giving some posterior samples is a good idea (we return to this below).
If you feel drawn to give the best sample,
  instead give the parameter values found by optimizing the posterior pdf,
  starting at the best sample as an initialization
  (and sacrifice your probabilism\note{something}).
This optimal-posterior parameter value is called the
  ``maximum \foreign{a posteriori}'' or ``MAP'' value for the parameters;
  it is like a maximum-likelihood value but regularized by the prior.
It is an estimator with some good (and some bad) properties, but it is not the point of \MCMC,
  and therefore outside the scope of this \documentname.

@@

There is one oddity to note here, as it catches many investigators by
surprise:
Even if the model is good and the posterior pdf is unimodal and well-behaved,
the median (or even mean) of the posterior pdf for each of parameters,
when taken together as a ``best-fit'' parameter vector $\pars^\ast$,
will not itself necessarily be a good fit to the data!
If there is substantial ``curvature'' to the pdf in the parameter
space, the mean or median of sampling does not necessarily lie at a
high-probability location of parameter space.
This may seem counterintuitive, but it is easy to see in the case of a
``banana-shaped'' posterior pdf.
This all relates to the fact that \MCMC\ is not an optimizer, it is a
sampler.
It is important, when reporting the output of an \MCMC\ run by giving
means or medians of the posterior pdf to remind the reader or user of
that output that it does not necessarily represent (collectively) a
good fit to the data; these outputs only give information that is
useful one parameter at a time.

@@

\paragraph{initialization-dependence:}
If you initialize your \MCMC\ runs in different places in parameter space,
  do you get the same (or very similar) final parameter samplings,
  in mean and variance?
If you don't,
  then your posterior pdf is badly multi-modal;
  trivial \MCMC\ methods probably won't fix your problem.
Not to put too fine a point on it:
If your results are initialization-dependent,
  then your \MCMC\ runs \emph{are not converged}.

You either have to go to a method that tries to explore the posterior
  more liberally, such as nested sampling or simulated tempering,
  or else split your ``model'' up into several sub-models,
  each of which contains different posterior modes.
In general, this problem is hard to fix;
  as we mentioned above, it is provably impossible
  to explore all of parameter space if the posterior pdf is
  complex in morphology.
We don't have very useful advice here, except to
  spend time learning about the multiple modes in the posterior pdf
  and what each ``means'' or corresponds to,
  and to do your best to express in your results the multiple optima.
One of the miracles of \MCMC\ is that if you \emph{can} fairly sample
  as complex posterior pdf,
  the fraction of samples in each mode
  tends to the relative total integrated posterior probability inside each mode,
  as the sampling converges.

@@

\paragraph{parameterization problems:}
It helps to think carefully about how to parameterize.
For example, if there is an angle $\phi$ in your problem,
  and your prior requires it to be between $0$ and $2\pi$,
  and the posterior mode is very near $0$,
  a little bit of probability leaking below $0$ plus the prior cutoffs plus mod-$2\pi$ symmetry
  makes an intrinsically unimodal problem multimodal!
In these cases, we usually advise reparameterization from a (say) amplitude $A$ and angle $\phi$
  to two vector components $a\equiv A\,\cos\phi$ and $b\equiv A\,\sin\phi$.
Of course this re-parameterization changes significantly the \emph{form} of the prior pdf
  (the transformation brings in a Jacobian);
  it has to be adjusted with care.
